**The status of the microbial census: an update**

### 2016-02-02 Schloss_Census2_mBio_2016.Rmd
Integrate Cameron's Thrashing of the Discussion :)

A census is typically carried out for people at a national level; however, microbial ecologists have implemented a molecular census of bacteria and archaea by sequencing their 16S rRNA genes. We assessed how well the microbial census of full-length 16S rRNA gene sequences is proceeding in the context of recent advances in high throughput sequencing technologies. Among the `r coverage_by_category_and_time["total","bact_count"]` and `r coverage_by_category_and_time["total","arch_count"]` full-length bacterial and archaeal sequences sequences, `r coverage_by_category_and_time["total","bact_good_coverage"]`% and `r coverage_by_category_and_time["total","arch_good_coverage"]`% of the bacterial and archeaeal sequences, respectively, belonged to operational taxonomic units (OTUs) that have been observed more than once. Although these metrics suggest that the census is approaching completion, `r coverage_by_category_and_time["total","bact_good_coverage"]`% of the bacterial and `r coverage_by_category_and_time["total","arch_good_coverage"]`% of the archaeal OTUs have been observed more than once. Thus, there is still considerable microbial diversity to be explored. Unfortunately, the number of new full-length sequences has been declining and these sequences are primarily deposited by a small number of studies. Furthermore, sequences from soil and aquatic environments, which are known to be rich in bacterial diversity, only represent `r coarse_percentages["S"]` and `r coarse_percentages["A"]`% of the sequences while sequences associated with zoonotic environments represent `r coarse_percentages["Z"]`% of the sequences. Continued use of traditional approaches and new technologies such as single cell genomics and short read assembly are likely to improve our ability to sample rare OTUs if it is possible to overcome this sampling biases. The success of ongoing efforts to use short read sequencing to characterize microbial communities requires that researchers strive to expand the depth and breadth of the microbial census.

### Importance

A census is a useful tool for assessing the characteristics of an entity, determining the best distribution of resources across individuals, and highlighting the strengths and weaknesses of the entity.

The biodiversity contained within the bacterial and archaeal domains dwarfs that of the eukaryotes and the services these organisms provide to the biosphere are critical. Surprisingly, we have done a relatively poor job of keeping track of the ongoing effort to characterize the biodiversity as represented in full-length 16S rRNA genes. By understanding how this census is proceeding, it is possible to suggest the best allocation of resources for advancing the census. We actually found that the ongoing effort has done an excellent job of sampling the most abundant organisms, but struggles to sample the more rare organisms. Through the use of new sequencing technologies we should be able to obtain full-length sequences from these rare organisms. Furthermore, we suggest that by allocating more resources to sampling environments known to have the greatest biodiversity we will be able to make significant advances in our characterization of microbial diversity.

@@

Considering the possibility that large culture-independent sequencing efforts may only be re-sequencing organisms that already exist in culture, we asked what percentage of OTUs had at least one cultured representative. We found that `r b_percent_cult_otu_all`% of the `r sum(b_cult_time[,'all_total_sobs'])` bacterial OTUs and `r a_percent_cult_otu_all`% of the `r sum(a_cult_time[,'all_total_sobs'])` archeael OTUs had at least one cultured representative (Figure 5). Comparing the percentage of sequences with cultured representatives to the percentage of OTUs containing a sequence from a cultured representative revealed strong cultivation biases for several phyla associated with model/biomedically relevant organisms (Figure 5). The Firmicutes and Proteobacteria both had higher percentages of sequences generated by cultivated representatives than would be expected based on the number of cultured organisms represented by OTUs. This likely reflects the extremely high number of cultivated *Escherichia coli*, *Bacillus*, *Streptococcus*, *Lactobacillus*, *Staphylococcus*, and others. Conversely, most phyla had a lower percentage of sequences belonging to cultivated representatives than would be expected based on the percentage of OTUs that have sequences from cultured organisms, indicating that the cultivation efforts in these clades are relatively efficient with regards to available diversity. It is clear that the majority of OTUs from any phylum remain uncultivated, to say nothing of the diversity of organisms that may be encapsulated within the 97% sequence identity cutoff.

@@

In spite of current estimates suggesting the global bacterial species richness may be as high as 10^12^ species [@Locey2015], the current census based on full-length 16S rRNA gene sequences suggests that our current sampling methods will prevent us from acquiring full-length sequences for that level of diversity. As we have shown, current strategies repeatedly sample the same OTUs and do a poor job of resampling rarer populations. Given this low level of OTU coverage, it is likely that there are many more bacterial and archaeal populations yet to be sampled. The recent advances in sequencing technologies have done little to expand the deposition of full-length sequences representing microbial biodiversity. We found that most sequences deposited into public databases are being made by a small number of projects that have deeply sampled similar environments. Furthermore, the number of full-length reads deposited into the databases has stalled. We fear that these trends will worsen unless researchers can leverage new sequencing and cultivation technologies to generate large numbers of full-length sequences from a large number of diverse samples.

During the period prior to the introduction of massively parallelized high throughput sequencing, it was common for a study to to generate dozens or hundreds of sequences per sample. The existing databases that are used for classifying sequences are based on these sequences. Thus, the sequences from those studies represent organisms that are generally abundant. We hypothesize that recent difficulties obtaining adequate classification for short sequences captured from more rare organisms are because our databases do not contain full-length references for those sequences. As stated in the 2004 census, we need to expand the diversity of environments that are deeply sequenced to improve OTU coverage. Efforts to census microbial life using short read technology such as the International Census of Marine Microbes, the Earth Microbiome Project, and the Human Microbiome Project have significantly advanced our knowledge of microbial biogeography; however, these analyses have demonstrated the limitations of databases and taxonomies that are based on sequences from common and abundant organisms.

It is clear that considerable biodiversity has been discovered since the first census in 2004. However, much of it has been biased towards particular phyla and environments. Our analysis suggests that `r b_goods`% of new full-length bacterial and archaeal sequences are likely to have already been seen. Meanwhile, `r b_schloss`% of bacterial and `r a_schloss`% of archaeal OTUs have only been observed once. It appears that although brute force deep sequencing of communities may increase our coverage of OTUs, new methods are needed to more directly sample rare organisms. Nevertheless, novel technologies such as single-cell genomics, metagenomics, and algorithms to recover full-length sequences from new sequencing platforms have demonstrated promise in circumventing previous limitations in identifying new OTUs. Using EMIRGE to assemble fragmented 16S rRNA gene amplicons may allow us to obtain deep coverage of communities; however, it is still unclear how faithful the assembled sequence is to that of the original organism. Additional sequencing technologies also offer the ability to directly generate full-length sequences, such as PacBio and potentially Oxford Nanopore. Initial application of PacBio to sequencing full-length fragments suggests that the sequences suffer from a high error rate. To obtain a more direct investigation of rare organisms, microbiologists are developing novel cultivation and single cell genomics techniques [@Nichols2010; @Buerger2012; @Thrash2014; @Das2015]. The ability to enrich or select for specific populations using these approaches could limit the need for redundant brute force sequencing. One strength of cultivation, single cell genomics, and metagenomics approaches is that not only do the methods yield a full 16S rRNA gene sequence, but they also allow researchers to obtain the complete genome of the organism. These approaches are still in active development, we hope that through continuous refinement, they may allow us to significantly improve the coverage of OTUs in public databases.

Our ability to assess the microbial census is limited by the quality of the data in the databases. First, it is widely acknowledged that the primers used to amplify 16S rRNA genes are biased; these biases are amplified when designing primers to amplify subregions used in sequencing short reads (Furhman, 2015). Furthermore, assembly of metagenomic data has shown the presence of introns in the 16S rRNA genes of organisms within the so-called "Candidate Phyla Radiation” (e.g. Saccharibacteria (TM7), Peregrinibacteria, Berkelbacteria (ACD58), WWE3 Microgenomates (OP11), Parcubacteria (OD1), et al.), that would preclude detection with standard PCR-based approaches [@Brown2015]. Thus, for many environments, the estimates for the census must be considered as lower bounds. Although the primers used to amplify the entire gene are also biased, obtaining a better characterization of microbial diversity will help improve the design of primers targeting subregions of the gene. Second, the willingness of researchers to contribute their sequences and the metadata describing the environment that the sequences were sampled from is critical for assessing the progress of the census and to accrue the benefits from having full-length sequences in the databases. Interestingly, the first 16S sequence was published in 1978, but was not available in a database until 1983. Similarly, only 5 of the 11 studies that used the EMIRGE algorithm deposited their sequences in GenBank. This makes the sequences from the other studies effectively invisible to the search algorithms used by 16S rRNA gene-specific databases to harvest sequences. As assembly and long read technologies advance, a mechanism is needed to assess the quality of the consensus sequences and to make them easily accessible to the 16S rRNA gene-specific databases.

The advance in sequencing technologies has significantly expanded our ability to peer into the depths of microbial communities. We now have unprecedented abilities to sequence large numbers of individual communities deeply for the same cost of generating hundreds of full length sequences. Yet, unless we invest in improving our full-length databases, it will continue to be difficult to contextualize this newly discovered biodiversity within the tree of life.

>>>>

A census is typically carried out for people at a national level; however, microbial ecologists have implemented a molecular census of bacteria and archaea by sequencing their 16S rRNA genes. We assessed how well the microbial census of full-length 16S rRNA gene sequences is proceeding in the context of recent advances in high throughput sequencing technologies. Among the `r coverage_by_category_and_time["total","bact_count"]` and `r coverage_by_category_and_time["total","arch_count"]` full-length bacterial and archaeal sequences sequences, `r coverage_by_category_and_time["total","bact_good_coverage"]`% and `r coverage_by_category_and_time["total","arch_good_coverage"]`% of the bacterial and archeaeal sequences, respectively, belonged to operational taxonomic units (OTUs) that have been observed more than once. Although these metrics suggest that the census is approaching completion, `r coverage_by_category_and_time["total","bact_schloss_coverage"]`% of the bacterial and `r coverage_by_category_and_time["total","arch_schloss_coverage"]`% of the archaeal OTUs have been observed more than once. Thus, there is still considerable microbial diversity to be explored. Unfortunately, the number of new full-length sequences has been declining and these sequences are primarily deposited by a small number of studies. Furthermore, sequences from soil and aquatic environments, which are known to be rich in bacterial diversity, only represent `r coarse_percentages["S"]` and `r coarse_percentages["A"]`% of the sequences while sequences associated with zoonotic environments represent `r coarse_percentages["Z"]`% of the sequences. Continued use of traditional approaches and new technologies such as single cell genomics and short read assembly are likely to improve our ability to sample rare OTUs if it is possible to overcome this sampling biases. The success of ongoing efforts to use short read sequencing to characterize microbial communities requires that researchers strive to expand the depth and breadth of the microbial census.

### Importance - too long

A census is a useful tool for assessing the characteristics of an entity, determining the best distribution of resources across individuals, and highlighting the strengths and weaknesses of the entity. The biodiversity contained within the bacterial and archaeal domains dwarfs that of the eukaryotes and the services these organisms provide to the biosphere are critical. Surprisingly, we have done a relatively poor job of keeping track of the ongoing effort to characterize the biodiversity as represented in full-length 16S rRNA genes. By understanding how this census is proceeding, it is possible to suggest the best allocation of resources for advancing the census. We actually found that the ongoing effort has done an excellent job of sampling the most abundant organisms, but struggles to sample the more rare organisms. Through the use of new sequencing technologies we should be able to obtain full-length sequences from these rare organisms. Furthermore, we suggest that by allocating more resources to sampling environments known to have the greatest biodiversity we will be able to make significant advances in our characterization of microbial diversity.

@@

Considering the possibility that large culture-independent sequencing efforts may only be re-sequencing organisms that already exist in culture, we asked what percentage of OTUs had at least one cultured representative. We found that `r b_percent_cult_otu_all`% of the `r sum(b_cult_time[,'all_total_sobs'])` bacterial OTUs and `r a_percent_cult_otu_all`% of the `r sum(a_cult_time[,'all_total_sobs'])` archeael OTUs had at least one cultured representative (Figure 5). Comparing the percentage of sequences with cultured representatives to the percentage of OTUs containing a sequence from a cultured representative revealed a strong cultivation bias within the Firmicutes, which had a higher percentage of sequences generated by cultivated representatives than would be expected based on the number of cultured organisms represented by OTUs (Figure 5). This likely reflects the extremely high number of cultivated biomedically relevant cultivars such as *Bacillus*, *Streptococcus*, *Lactobacillus*, *Staphylococcus*, and others. Conversely, many phyla, including Cyanobacteria, Actinobacteria, Bacteroidetes, and Nitrospirae, had a lower percentage of sequences belonging to cultivated representatives than would be expected based on the percentage of OTUs that have sequences from cultured organisms, indicating that the cultivation efforts in these clades are relatively efficient with regards to available diversity. Nevertheless, it is clear that the majority of OTUs from any phylum remain uncultivated, to say nothing of the diversity of organisms that may be encapsulated within the 97% sequence identity cutoff.

@@

It is clear that considerable biodiversity has been discovered since the first census in 2004. However, much of it has been biased towards particular phyla and environments. Our analysis suggests that `r b_goods`% of new full-length bacterial and archaeal sequences are likely to have already been seen. Meanwhile, `r b_schloss`% of bacterial and `r a_schloss`% of archaeal OTUs have only been observed once. In spite of current estimates suggesting the global bacterial species richness may be as high as 10^12^ species [@Locey2015], the current census based on full-length 16S rRNA gene sequences suggests that existing sampling methods will prevent us from acquiring full-length sequences for that level of diversity. As we have shown, current strategies repeatedly sample the same OTUs and do a poor job of resampling rarer populations. Given this low level of OTU coverage, it is likely that there are many more bacterial and archaeal populations yet to be sampled.

There are several additional reasons to suspect that the current census should be considered conservative. First, we found that most sequences recently deposited into public databases are being made by a small number of projects that have deeply sampled similar environments, and the number of full-length reads deposited into the databases has stalled. Second, it is widely acknowledged that 16S rRNA gene primers are biased; these biases are amplified when designing primers to amplify subregions used in sequencing short reads [@Parada2015]. Assembly of metagenomic data has shown the presence of introns in the 16S rRNA genes of organisms within the so-called "Candidate Phyla Radiation” (e.g. Saccharibacteria (TM7), Peregrinibacteria, Berkelbacteria (ACD58), WWE3 Microgenomates (OP11), Parcubacteria (OD1), et al.) that would preclude detection with standard PCR-based approaches [@Brown2015; @EloeFadrosh2016]. Third, the willingness of researchers to contribute their sequences and the metadata describing the environment that the sequences were sampled from is critical for assessing the progress of the census and to accrue the benefits from having full-length sequences in the databases. Interestingly, the first 16S sequence was published in 1978, but was not available in a database until 1983. Similarly, only 5 of the 11 studies that used the EMIRGE algorithm deposited their sequences in GenBank. This makes the sequences from the other studies effectively invisible to the search algorithms used by 16S rRNA gene-specific databases to harvest sequences. As assembly and long read technologies advance, a mechanism is needed to assess the quality of the consensus sequences and to make them easily accessible to the 16S rRNA gene-specific databases.

Efforts to census microbial life using short read technology such as the International Census of Marine Microbes, the Earth Microbiome Project, and the Human Microbiome Project have significantly advanced our knowledge of microbial biogeography; however, these analyses have demonstrated the limitations of databases and taxonomies that are based on sequences from common and abundant organisms. During the period prior to the introduction of massively parallelized high throughput sequencing, it was common for a study to generate dozens or hundreds of sequences per sample. The existing databases that are used for classifying sequences are based on those sequences, which represent organisms that are generally abundant. We hypothesize that recent difficulties obtaining adequate classification for short sequences captured from more rare organisms are because our databases do not contain full-length references for those sequences. We fear that these trends will worsen unless researchers can leverage new sequencing and cultivation technologies to generate large numbers of full-length sequences from a large number of diverse samples.

Novel technologies such as single-cell genomics, metagenomics, and algorithms to recover full-length sequences from new sequencing platforms have demonstrated promise in circumventing previous limitations in identifying new OTUs. Using EMIRGE to assemble fragmented 16S rRNA gene amplicons may allow us to obtain deep coverage of communities; however, it is still unclear how faithful the assembled sequence is to that of the original organism. Additional sequencing technologies also offer the ability to directly generate full-length sequences, such as PacBio and potentially Oxford Nanopore. Initial application of PacBio to sequencing full-length fragments suggests that the sequences suffer from a high error rate [@Schloss2015]. To obtain a more direct investigation of rare organisms, microbiologists are developing novel cultivation and single cell genomics techniques [@Nichols2010; @Buerger2012; @Thrash2015; @Das2015]. The ability to enrich or select for specific populations using these approaches could limit the need for redundant brute force sequencing. These approaches are still in active development, and we hope that through continuous refinement, they may allow us to significantly improve the coverage of OTUs in public databases.
